Single-seismometer Focal Mechanism and Uncertainty Estimation from Body Waves

### 2024-12-10 project/paper/paper.tex
organize

    Combining equations \ref{eq:7} and \ref{eq:8} gives the plane containing every point of 
    tangency:
    \begin{align} \label{eq:9}
      \frac{1}{2} (\nabla_{\bm{A}} E)^\top \bm{A} - f(\bm{A}) = 
              \bm{A}_o^\top \tilde{\bm{\Sigma}}^{-1} (\bm{A} - \bm{A}_o)  = -1
    \end{align}

@@

    To find the constrained points of tangency, we can intersect the original ellipsoid with planes
    \ref{eq:9} and \ref{eq:10}. The line of intersection is such that
    \begin{align} \label{eq:11}
      \bm{A}(t) = (1 - \nicefrac{1}{c_1}) \bm{A}_o + t\bm{D}
    \end{align}

@@

    \noindent where $c_2 = \bm{A}_o^\top \tilde{\bm{\Sigma}}^{-1} \bm{A}_s = \langle\bm{A}_o^\top, \bm{A}_s\rangle_{\tilde{\bm{\Sigma}}^{-1}}$.

    Combining equations \ref{eq:7} and \ref{eq:11} produces a quadratic parameterized by the scalar $t$,
    from which we can obtain a quadratic equation for the points of tangency:

@@

    Denote $c_3 = \bm{A}_s^\top \tilde{\bm{\Sigma}}^{-1} \bm{A}_s = \|\bm{A}_s\|_{\tilde{\bm{\Sigma}}^{-1}}^2$. Then by
    substitution we have
    ${\bm{D}^\top \tilde{\bm{\Sigma}}^{-1} \bm{D}} = c_1(c_1c_3 - c_2^2)$ and ${\bm{A}_o^\top \tilde{\bm{\Sigma}}^{-1} \bm{D}} = 0$.
    Solving then yields the necessary points (denoted $\bm{A}_b$) as
    \begin{align} \label{eq:15}
      \bm{A}_b = \left(1 - \frac{1}{c_1}\right)\bm{A}_o \pm
        \left(\frac{1}{c_1} \sqrt{\frac{c_1 - 1}{c_1c_3 - c_2^2}}\right)(c_1\bm{A}_s - c_2\bm{A}_o)
    \end{align}

@@

    Thus, given 2 nonparallel vectors $\bm{A}_o$ and $\bm{A}_s$ and positive definite covariance
    matrix $\tilde{\bm{\Sigma}}$, we accept the simulation if $\hat{\bm{A}}_o^\top \hat{\bm{A}_s} \geq \text{tol} 
    = \min_b \hat{\bm{A}}_o^\top \hat{\bm{A}_b}$. The condition, more formally, is
    \begin{align} \label{eq:16}
      \max_b \hat{\bm{A}}_o^\top (\hat{\bm{A}}_b - \hat{\bm{A}}_s) \leq 0
    \end{align}

    \noindent where $\hat{\bm{A}}_o$, $\hat{\bm{A}}_s$ and $\hat{\bm{A}}_b$ are unit vectors.

@@

  \subsection{Misfit gradient}
    The vector $\bm{A} = (A^P, A^{SV}, A^{SH})^\top$ can be treated as a continuously differentiable 
    nonlinear function of parameters $\bm{m} = (\psi, \delta, \lambda)^\top$, defined by equation 
    \ref{eq:1}. We can thus apply directed search methods to minimize the following cost function:
    \begin{align} \label{eq:17}
      \Phi(\bm{m}) = -\hat{\bm{A}}_o^\top \hat{\bm{A}}
    \end{align}

@@

    Define a vector $\bm{\eta} = (s_R, q_R, p_R, q_L, p_L)^\top$. Then we can encode
    the misfit as
    \begin{align} \label{eq:18}
      \Phi(\bm{m}) = \tilde{\Phi}(\bm{A}(\bm{\eta}(\bm{m})))
    \end{align}

@@

    i) From $\bm{m}$ to $\bm{\eta}$, we can use equation \ref{eq:2} to define the Jacobian
    as follows:
    \begin{align*}
      &a = \sin(\lambda)\;\;\;\;\;\;\;\;\;\;\;\;\; b = \cos(2\delta)\;\;\;\;\;\;\;\;\;\;\; c = \cos(\lambda)\\
      &d = \sin(2\delta)\;\;\;\;\;\;\;\;\;\;\;\; e = \sin(\delta)\;\;\;\;\;\;\;\;\;\;\;\; f = \cos(\delta)\\
      &g = \sin(\psi - \phi)\;\;\;\;\;\; h = \cos(\psi - \phi)\;\;\;\;\; p = cos(2(\psi - \phi))\\
      &q = \sin(2(\psi - \phi))
    \end{align*}
    \begin{align} \label{eq:19}
      \nonumber&\bm{J_p}(\bm{\eta}) = \\
      &\begin{bmatrix}
        0         &ab         &0.5cd\\
        abh-cfg   &-2adg-ceh  &cbg-afh\\
        2cep+adq  &cfq-abp    &-aeq-0.5cdp\\
        -cfh-abg  &ceg-2adh   &afg+cbh\\
        adp-2ceq  &abq+cfp    &0.5cdq-aep
      \end{bmatrix}
    \end{align}

    ii) From $\bm{\eta}$ to $A$, we have a linear transformation whose Jacobian is the matrix
    of coefficients from equation \ref{eq:1}:
    \begin{align} \label{eq:20}
      \nonumber&\bm{J_\eta}(\bm{A}) = \\
      &\begin{bmatrix}
        \frac{3\cos^2(i)-1}{\alpha_h^3}    &\frac{-\sin(2i)}{\alpha_h^3}    &\frac{-sin^2(i)}{\alpha_h^3}   &0    &0\\
        \frac{1.5\sin(2j)}{\beta_h^3}      &\frac{\cos(2j)}{\beta_h^3}      &\frac{0.5\sin(2j)}{\beta_h^3}  &0    &0\\
        0         &0           &0        &\frac{\cos(j)}{\beta_h^3}       &\frac{\sin(j)}{\beta_h^3}
      \end{bmatrix}
    \end{align}

    iii) From $\bm{A}$ to $\Phi$, we have a scalar function whose Jacobian is the gradient
    transposed.
    \begin{align} \label{eq:21}
      \nonumber\nabla_{\bm{A}}\tilde{\Phi} &= \frac{\|\bm{A}\|\nabla_{\bm{A}}(-\hat{\bm{A}}_o^\top\bm{A}) - 
      (-\hat{\bm{A}}_o^\top \bm{A}) \nabla_{\bm{A}}(\|\bm{A}\|)}{\|\bm{A}\|^2}\\
            &= \frac{-1}{\|\bm{A}\|}(\hat{\bm{A}}_o + \Phi\hat{\bm{A}})
    \end{align}

    Combining \ref{eq:19}, \ref{eq:20} and \ref{eq:21} gives the overall gradient:
    \begin{align} \label{eq:22}
      \nabla_{\bm{m}}\Phi = \bm{J_p}(\bm{\eta})^\top \bm{J_\eta}(\bm{A})^\top \nabla_{\bm{A}}\tilde{\Phi}
    \end{align}

  \subsection{Optimization algorithms}
    \begin{figure}[ht!]
      \includegraphics[width=8.6cm]{seed_2024} 
      \caption{Results of steepest descent algorithm after a systematic sweep of starting points.}
      \label{fig:2}
    \end{figure}

@@

    \subsection{Kagan angle distribution}
      An alternative way of quantifying model deviations.

>>>>

    \noindent Combining equations \ref{eq:7} and \ref{eq:8} gives the plane containing every point of 
    tangency:
    \begin{align} \label{eq:9}
      \frac{1}{2} (\nabla_{\bm{A}} E)^\top \bm{A} - f(\bm{A}) = 
              \bm{A}_o^\top \tilde{\bm{\Sigma}}^{-1} (\bm{A} - \bm{A}_o)  = -1
    \end{align}

@@

    \noindent To find the constrained points of tangency, we can intersect the original ellipsoid with planes
    \ref{eq:9} and \ref{eq:10}. The line of intersection is such that
    \begin{align} \label{eq:11}
      \bm{A}(t) = (1 - \nicefrac{1}{c_1}) \bm{A}_o + t\bm{D}
    \end{align}

@@

    \noindent where $c_2 = \bm{A}_o^\top \tilde{\bm{\Sigma}}^{-1} \bm{A}_s = \langle\bm{A}_o^\top, \bm{A}_s\rangle_{\tilde{\bm{\Sigma}}^{-1}}$.
    Equation \ref{eq:13} is the triple product expansion of constituent vectors.

    \noindent Combining equations \ref{eq:7} and \ref{eq:11} produces a quadratic parameterized by the scalar $t$,
    from which we can obtain a quadratic equation for the points of tangency:

@@

    \noindent Denote $c_3 = \bm{A}_s^\top \tilde{\bm{\Sigma}}^{-1} \bm{A}_s = \|\bm{A}_s\|_{\tilde{\bm{\Sigma}}^{-1}}^2$. By
    substitution we have
    ${\bm{D}^\top \tilde{\bm{\Sigma}}^{-1} \bm{D}} = c_1(c_1c_3 - c_2^2)$ and ${\bm{A}_o^\top \tilde{\bm{\Sigma}}^{-1} \bm{D}} = 0$.
    Solving then yields the necessary points (denoted $\bm{A}_b$) as
    \begin{align} \label{eq:15}
      \bm{A}_b = \left(1 - \frac{1}{c_1}\right)\bm{A}_o \pm
        \left(\frac{1}{c_1} \sqrt{\frac{c_1 - 1}{c_1c_3 - c_2^2}}\right)(c_1\bm{A}_s - c_2\bm{A}_o)
    \end{align}

@@

    \noindent The calculated boundary points are necessary to design reliable Monte-Carlo methods for uncertainty
    quantification (see section \ref{sec:uncertainty}).

    % Thus, given 2 nonparallel vectors $\bm{A}_o$ and $\bm{A}_s$ and positive definite covariance
    % matrix $\tilde{\bm{\Sigma}}$, we accept the simulation if $\hat{\bm{A}}_o^\top \hat{\bm{A}_s} \geq \text{tol} 
    % = \min_b \hat{\bm{A}}_o^\top \hat{\bm{A}_b}$. The condition, more formally, is
    % \begin{align} \label{eq:16}
    %   \max_b \hat{\bm{A}}_o^\top (\hat{\bm{A}}_b - \hat{\bm{A}}_s) \leq 0
    % \end{align}

    % \noindent where $\hat{\bm{A}}_o$, $\hat{\bm{A}}_s$ and $\hat{\bm{A}}_b$ are unit vectors.

@@

  \subsection{Misfit gradient}
    The vector $\bm{A} = (A^P, A^{SV}, A^{SH})^\top$ can be treated as a continuously differentiable 
    nonlinear function of parameters $\bm{m} = (\psi, \delta, \lambda)^\top$, defined by equation 
    \ref{eq:1}. We can thus apply directed search methods to minimize the following cost function:
    \begin{align} \label{eq:16}
      \Phi(\bm{m}) = -\hat{\bm{A}}_o^\top \hat{\bm{A}}
    \end{align}

@@

    Define a vector $\bm{\eta} = (s_R, q_R, p_R, q_L, p_L)^\top$. Then we can encode
    the misfit as
    \begin{align} \label{eq:17}
      \Phi(\bm{m}) = \tilde{\Phi}(\bm{A}(\bm{\eta}(\bm{m})))
    \end{align}

@@

    i) From $\bm{m}$ to $\bm{\eta}$, we can use equation \ref{eq:2} to define the Jacobian
    as follows:
    \begin{align*}
      &a = \sin(\lambda)\;\;\;\;\;\;\;\;\;\;\;\;\; b = \cos(2\delta)\;\;\;\;\;\;\;\;\;\;\; c = \cos(\lambda)\\
      &d = \sin(2\delta)\;\;\;\;\;\;\;\;\;\;\;\; e = \sin(\delta)\;\;\;\;\;\;\;\;\;\;\;\; f = \cos(\delta)\\
      &g = \sin(\psi - \phi)\;\;\;\;\;\; h = \cos(\psi - \phi)\;\;\;\;\; p = cos(2(\psi - \phi))\\
      &q = \sin(2(\psi - \phi))
    \end{align*}
    \begin{align} \label{eq:18}
      \nonumber&\bm{J_p}(\bm{\eta}) = \\
      &\begin{bmatrix}
        0         &ab         &0.5cd\\
        abh-cfg   &-2adg-ceh  &cbg-afh\\
        2cep+adq  &cfq-abp    &-aeq-0.5cdp\\
        -cfh-abg  &ceg-2adh   &afg+cbh\\
        adp-2ceq  &abq+cfp    &0.5cdq-aep
      \end{bmatrix}
    \end{align}

    ii) From $\bm{\eta}$ to $A$, we have a linear transformation whose Jacobian is the matrix
    of coefficients from equation \ref{eq:1}:
    \begin{align} \label{eq:19}
      \nonumber&\bm{J_\eta}(\bm{A}) = \\
      &\begin{bmatrix}
        \frac{3\cos^2(i)-1}{\alpha_h^3}    &\frac{-\sin(2i)}{\alpha_h^3}    &\frac{-sin^2(i)}{\alpha_h^3}   &0    &0\\
        \frac{1.5\sin(2j)}{\beta_h^3}      &\frac{\cos(2j)}{\beta_h^3}      &\frac{0.5\sin(2j)}{\beta_h^3}  &0    &0\\
        0         &0           &0        &\frac{\cos(j)}{\beta_h^3}       &\frac{\sin(j)}{\beta_h^3}
      \end{bmatrix}
    \end{align}

    iii) From $\bm{A}$ to $\Phi$, we have a scalar function whose Jacobian is the gradient
    transposed.
    \begin{align} \label{eq:20}
      \nonumber\nabla_{\bm{A}}\tilde{\Phi} &= \frac{\|\bm{A}\|\nabla_{\bm{A}}(-\hat{\bm{A}}_o^\top\bm{A}) - 
      (-\hat{\bm{A}}_o^\top \bm{A}) \nabla_{\bm{A}}(\|\bm{A}\|)}{\|\bm{A}\|^2}\\
            &= \frac{-1}{\|\bm{A}\|}(\hat{\bm{A}}_o + \Phi\hat{\bm{A}})
    \end{align}

    Combining \ref{eq:18}, \ref{eq:19} and \ref{eq:20} gives the overall gradient:
    \begin{align} \label{eq:21}
      \nabla_{\bm{m}}\Phi = \bm{J_p}(\bm{\eta})^\top \bm{J_\eta}(\bm{A})^\top \nabla_{\bm{A}}\tilde{\Phi}
    \end{align}

  \subsection{Optimization algorithms}

  Insert description here to be backed up by figure 2...

  Why isn't this doing its thing?

  I know you can do it

  Just a few more paragraphs

    \begin{figure}[ht!]
      \includegraphics[width=8.6cm]{seed_2024} 
      \caption{Results of steepest descent algorithm for randomly sampled starting points.}
      \label{fig:2}
    \end{figure}

@@
